---
title: "Class 11: Structural Bioinformatics pt2"
author: "Maria Tavares"
format: pdf
---

##Background

We saw last day that PDB has 209,886 entries (Oct/Nov 2025). UniProtKB (i.e. protain sequence database) has 199,579,901 entries. 

```{r}
209886/199579901 * 100
```

So the PDB has only 0.1% coverage of the main sequence database. 

Enter AlphaFold data base (AFDB) < https://alphafold.ebi.ac.uk > that attempts to provide computed models for all sequences in UniProt. 

"AlphaFold DB provides open access to over 200 million protein structure predictions to accelerate scientific research"

## AlphaFold

AlphaFold has 3 main outputs 

- the predicted coordinates (PDB files)
- A local quality score called **plDDT** (one for each amino acid)
- A second quality score **PAE** Predicted Alignes Error (for each pair of amino-acid)

We can run AlphaFold ourselves if we not happy with AFDB (i.e. no coverage or poor model)

## Interpreting/analyzing AF results in R 

```{r}
results_dir <- "HIVPR_dimer_23119/"
```

```{r}
# File names for all PDB models
pdb_files <- list.files(path=results_dir,
                        pattern="*.pdb",
                        full.names = TRUE)

# Print our PDB file names
basename(pdb_files)
```

```{r}
library(bio3d)

# Read all data from Models 
#  and superpose/fit coords
pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")
```

```{r}
pdbs
```



