---
title: "Class05: data viz with ggplot"
author: "Maria (PID A69036242)"
format: pdf
---

Today we are playing with plotting and graphics in R. 

There are lots of ways to make cool figures in R. 
There is "base" R graphics (`plot()`, `hist()`, `boxplot()` etc)

There is also add-on packages, like **ggplot**.

```{r}
head(cars)
```

Let's plot this with "base" R: 

```{r}
plot(cars)
```

```{r}
head(mtcars)
```

Let's plot `mpg` vs `disp`

```{r}
plot(mtcars$mpg, mtcars$disp, pch=16)
```

```{r}
hist(mtcars$mpg)
```

## GGPLOT 

The main function in the ggplot2 package is `ggplot()`. First I need to install the **ggplot2** package. I can install any package with the function `install.packages()`

> **N.M** I never want to run `install.packages()` in my quarto source document!! 

```{r}
library(ggplot2)
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()
```
```{r}
ggplot(cars) +
  aes(speed) +
  geom_histogram()
```

Every ggplot needs at least 3 things:

- The **data** (given with `ggplot(cars)`)
- The **aesthetic** mapping (given the `aes()`)
- The **geom** (given by `geom_poin()`)

> For simple canned graphs "base" R is nearly always faster

### Adding more layers

Let's add a line and a title, subtitle and caption as well as a custom axis labels 

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_line()
```

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth()
```

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE)
```

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  labs(title = "Silly Plot", x = "Speed (MPH)", y = "Distance(ft)", caption = "when is time anyways?") + 
  theme_bw()

```

Read data file from online URL
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> Q1. How many genes are in this dataset? 

There are `r nrow(genes)` in this dataset 

> Q2. How many "up" regulated genes are there?

```{r}
sum(genes$State == "up")
ncol(genes)
```

There are `r sum(genes$State == "up")` "up" genes

```{r}
table(genes$State)
```

```{r}
ggplot(genes) +
  aes(x = Condition1, y = Condition2) +
  geom_point(col="blue")
```

```{r}
p <- ggplot(genes) +
  aes(x = Condition1, y = Condition2, col = State) +
  geom_point() +
  scale_color_manual(values=c("blue", "gray", "red"))
```

```{r}
p + labs(title="look at me")
```

Silly example of adding labels

```{r}
ggplot(genes) +
  aes(x = Condition1,
      y = Condition2, 
      col = State, 
      label=Gene) +
  geom_point() +
  scale_color_manual(values=c("blue", "gray", "red")) +
  geom_text()
```

```{r}
library(ggrepel)
ggplot(genes) +
  aes(x = Condition1, y = Condition2, col = State, label = Gene) +
  geom_point() +
  scale_color_manual(values=c("blue", "gray", "red")) +
  geom_text_repel(max.overlaps = 60) +
  theme_bw()

```

## Going Further

Playing with some different layers and the gapminder dataset... 

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder)
```

```{r}
tail(gapminder)
```

```{r}
ggplot(gapminder) +
  aes(x = gdpPercap,
      y = lifeExp, 
      col=continent,
      size = pop) +
  geom_point(alpha=0.4) +
  facet_wrap(~continent) +
  theme_bw()
```



