---
title: "Class 17: Genome Informatics"
author: Maria Tavares
format: pdf
---

```{r}
library(tidyverse)
```

```{r}
library(ggplot2)
library(readr)
```

```{r}
galaxy_data <- read.table("rs8067378_ENSG00000172057.6.txt")
```

> Q13: Read this file into R and determine the sample size for each genotype and
their corresponding median expression levels for each of these genotypes.

```{r}
galaxy_data |>
  group_by(geno) |>
  summarize(
    Sample_Size = n(),
    Median_Expression = median(exp)
  )
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

ORMDL3 expression is highest in individuals with the A/A genotype and lowest in those with the G/G genotype. This suggests that the SNP affects ORMDL3 expression, with the G allele associated with reduced expression in a dose-dependent manner.

```{r}
ggplot(galaxy_data) +
  geom_boxplot(aes(x = geno, y = exp, fill = geno, notch=TRUE)) +
  labs(
    x = "Genotype",
    y = "Expression Level"
 
)+ theme_minimal()
```



