---
title: "Class 10: Halloween Mini Project"
author: "Maria Tavares (PID A69036242)"
format: pdf
---

## Background

It's Halloween tonight and to get in the mood let's explore some candy data.

## Data Import

```{r}
candy_file <- "candy-data.csv"

candy <- read.csv(candy_file, row.names=1)
head(candy)
```

```{r}
flextable::flextable(head(candy))
```

## Favorite Candy 

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

```{r}
candy["Twix", ]$winpercent
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Reese's Peanut Butter cup", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
library("skimr")
skim(candy)
```

```{r, message = FALSE}
library(dplyr)

candy |> nrow()

candy |> select(winpercent)
```

```{r}
win <- candy$winpercent
win.mean <- mean(win)
round(win.mean)
```

```{r}
candy%>% select (winpercent)
candy |> 
  select(winpercent)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

`Winpercent` is 0 to 100 percent scale. 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

Zero means it's not chocolate and one means it is chocolate. 

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)
ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins = 20)
```

> Q9. Is the distribution of winpercent values symmetrical?

No, it is skewed.

```{r}
ggplot(candy) +
  aes(winpercent) +
  geom_density()
```

> Q10. Is the center of the distribution above or below 50%?

Below. 
```{r}
mean(candy$winpercent)
summary(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# 1. Find the chocolate candy in the dataset
choc.inds <- as.logical(candy$chocolate) 
choc.inds
choc.candy <- candy[choc.inds,]

# 2. Extract their meanpercent values 
choc.win <- choc.candy$winpercent

# 3. Find the mean of these values 
mean (choc.win)

# 4-6. Do the same for fruity candy
fruity.inds <- as.logical(candy$fruity) 
fruity.inds
fruity.candy <- candy[fruity.inds,]
fruity.win <- fruity.candy$winpercent
mean(fruity.win)

# 7. Which mean value is higher 
mean(choc.win) > mean(fruity.win)

```


> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruity.win)
```

## Overall Candy Rating

> Q13. What are the five least liked candy types in this set?

```{r}
candy %>% arrange(winpercent) %>% head(5)
```
```{r}
ord.ind <- order(candy$winpercent)
head(candy[order(candy$winpercent),], n=5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>% arrange(winpercent) %>% tail(5)
```
```{r}
tail(candy[order(candy$winpercent),], n=5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}

ggplot(candy) +
  aes(winpercent, rownames(candy), winpercent) +
  geom_col() 

```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col() 
```

Add some color 

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"


ggplot(candy) +
  aes(x=winpercent, 
      y=reorder(rownames(candy), winpercent)) +
  geom_col(fill=my_cols) 
```

Now, for the first time, using this plot we can answer questions like:
> Q17. What is the worst ranked chocolate candy? 

Nik L Nip

> Q18. What is the best ranked fruity candy?

Reese's Peanut Butter Cup

## Price percent 

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 10)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

## Correlation

```{r}
cij <- cor(candy)
```

```{r}
library(corrplot)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

chocolate & fruity

> Q23. Similarly, what two variables are most positively correlated?

chocolate & winpercent 

## PCA

Let’s apply PCA using the `prcom()` function to our candy dataset remembering to set the `scale=TRUE` argument

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label = rownames(pca$x)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols) +
  labs(title="Barry's candy space map")
```

Don't forget about your variable "loadings" - how the original variables contribute to your new pcs

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation))+
  geom_col()
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Pluribus, hard, and fruity. It makes sense because the fruity candles usually are harder than chocolate and come in packages with multiple (like skittles, starburst).

